---  
title: Continuous Probability Distributions
output: 
  html_document
---
  
## Some continuous probability distributions


### Uniform Distribution

* valid between a minimum and a maximum value
* all intervals _of the same length_ between the minimum and the maximum values are equally probable

$$ P(X = x; a,b) = \frac{1}{b - a}, x \in [a,b]$$

* application: often used as a prior distribution when we have no prior knowledge
* examples for discrete version of uniform distribution: die rolls
* mean: $\frac{1}{2} (a+b)$
* variance: $\frac{1}{12}(b-a)^2$

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

n <- 10
k <- seq(0,10,0.01)
minval = 3
maxval = 7
ggplot(data_frame(k), aes(k, dunif(k, minval, maxval))) + geom_line(color = "blue") + ylab("p = uniform(3,7)") +
  ggtitle("Uniform distribution example")
ggplot(data_frame(k), aes(k, punif(k, minval, maxval))) + geom_line(color = "blue") + ylab("p = uniform(3,7)") +
  ggtitle("Corresponding cumulative distribution")
```


### Normal / Gaussian Distribution

#### General

* the famous "bell curve"
* characterized by mean $\mu$ and variance $\sigma^2$:

$$ P(X = x; \mu, \sigma^2) = (2\pi \sigma^2)^{-1/2}e^{-(x - \mu)^2/2\sigma^2}$$

* mean: $\mu$
* variance: $\sigma^2$

* example: measurement errors (if not systematic)

Compare normal distributions with different variances:

```{r, warning=FALSE, message=FALSE}
library(ggfortify)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'blue', mean=0, sd=1)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'green', p = p, mean=0, sd=0.5)
ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'cyan', p = p, mean=0, sd=2)
```

... and their corresponding CDFs:

```{r, warning=FALSE, message=FALSE}
p <- ggdistribution(pnorm, seq(-4, 4, 0.1), colour = 'blue', mean=0, sd=1)
p <- ggdistribution(pnorm, seq(-4, 4, 0.1), colour = 'green', p = p, mean=0, sd=0.5)
ggdistribution(pnorm, seq(-4, 4, 0.1), colour = 'cyan', p = p, mean=0, sd=2)
```

#### Why is the Gaussian so important? 

* many processes in nature follow a Gaussian distribution, because they _add up many many small fluctuations_
* for a given mean and variance, the Gaussian is the maximum entropy (least informative) distribution, the one that can come about in the largest number of ways

Adding up small fluctuations: example

* Say you're standing on a road, together with 10 other people, and repeatedly flip a coin. If it turns up heads, you turn left, if it turns up tails, you turn right. (The other people do the same, flipping coins and turning either left or right.) Say all of you flip the coin 10 times. How will the group of you be distributed on the road?
* Now say you are not 11 people, but 111. How will the distribution look now?
* Finally, how about if there were 1111 people of you doing this?

```{r}
xs <- lapply(c(11,111,1111), function(x) {replicate(x, sum(sample(c(-1,1), 10, replace = TRUE)))})
plots <- lapply(xs, function(x) {ggplot(data_frame(x = x), aes(x)) + geom_histogram(binwidth = 1)})
do.call('grid.arrange', list('grobs' = plots, 'ncol' = 3))
```


The larger the number of small fluctuations included, the more normal the distribution becomes.

&nbsp;

#### The standard normal distribution

If $\mu = 0, \sigma^2 =1$, we have a _standard normal_ distribution.

```{r}
x <- seq(-4, 4, 0.1)
#y <- dnorm(x,0,1)
ggplot(data_frame(x), aes(x)) + stat_function(fun = dnorm, args = list(mean = 0, sd =1))
```

The standard normal is tremendously important in statistics. 

* ~ 68% of the standard normal density lies between $[\bar{x} - 1 sd, \bar{x} + 1 sd]$
* ~ 95% of the standard normal density lies between $[\bar{x} - 2 sd, \bar{x} + 2 sd]$
* ~ 99% of the standard normal density lies between $[\bar{x} - 3 sd, \bar{x} + 3 sd]$

```{r}
normal_prob_area_plot <- function(lb, ub, mean = 0, sd = 1, limits = c(mean - 4 * sd, mean + 4 * sd), fcolor = "blue") {
    x <- seq(limits[1], limits[2], length.out = 100)
    xmin <- max(lb, limits[1])
    xmax <- min(ub, limits[2])
    areax <- seq(xmin, xmax, length.out = 100)
    area <- data.frame(x = areax, ymin = 0, ymax = dnorm(areax, mean = mean, sd = sd))
    (ggplot()
     + geom_line(data.frame(x = x, y = dnorm(x, mean = mean, sd = sd)),
                 mapping = aes(x = x, y = y))
     + geom_ribbon(data = area, mapping = aes(x = x, ymin = ymin, ymax = ymax), fill = fcolor)
     + scale_x_continuous(limits = limits))
}
g1 <- normal_prob_area_plot(-3,3)
g2 <- normal_prob_area_plot(-2,2, fcolor = "white")
g3 <- normal_prob_area_plot(-1,1, fcolor = "green")
grid.arrange(g3,g2,g1, ncol=2, nrow=2)
```


#### Converting between standardized and non-standardized normals

* to go from normal to standard normal:
    + normal: $$X \sim N(\mu,\sigma^2)$$
    + standard normal: $$Z = \frac{X -\mu}{\sigma} \sim N(0, 1)$$

* to go from standard normal to normal:
    + standard normal: $$Z \sim N(0, 1)$$
    + normal: $$X = \mu + \sigma Z \sim N(\mu, \sigma^2)$$



