---  
title: Continuous Probability Distributions
output: 
  html_document
---
  
## Some continuous probability distributions


### Uniform Distribution

* valid between a minimum and a maximum value
* all intervals _of the same length_ between the minimum and the maximum values are equally probable

$$ P(X = x; a,b) = \frac{1}{b - a}, x \in [a,b]$$

* application: often used as a prior distribution when we have no prior knowledge
* examples for discrete version of uniform distribution: die rolls
* mean: $\frac{1}{2} (a+b)$
* variance: $\frac{1}{12}(b-a)^2$

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

n <- 10
k <- seq(0,10,0.01)
minval = 3
maxval = 7
prob = dunif(k, minval, maxval)
ggplot(data_frame(x = k, y = prob), aes(x, y)) + geom_line(color = "blue") + ylab("p = uniform(3,7)") +
  ggtitle("Uniform distribution example")
```


### Normal / Gaussian Distribution

* the famous "bell curve"
* characterized by mean $\mu$ and variance $\sigma^2$:

$$ P(X = x; \mu, \sigma^2) = (2\pi \sigma^2)^{-1/2}e^{-(x - \mu)^2/2\sigma^2}$$
* mean: $\mu$
* variance: $\sigma^2$

* examples: 

Compare normal distributions with different variances:

```{r, warning=FALSE, message=FALSE}
library(ggfortify)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'blue', mean=0, sd=1)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'green', p = p, mean=0, sd=0.5)
ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'cyan', p = p, mean=0, sd=2)
```

Why are so many processes / variables normally distributed?


If $\mu = 0, \sigma^2 =1$, we have a _standard normal_ istribution.

```{r}
x <- seq(-4, 4, 0.1)
y <- dnorm(x,0,1)
ggplot(data_frame(x), aes(x)) + stat_function(fun = dnorm, args = list(mean = 0, sd =1))
```

The standard normal is tremendously important in statistics, as 
