---  
title: Discrete Probability Distributions
output: 
  html_document
---
  
  
## Some discrete probability distributions


### Bernoulli Distribution

* outcome is binary with values 0 or 1
* parameter(s): p = probability of success (1)
* examples: coin flips, rain vs. no rain, ...
* distribution:

$$P(X = x; p) =  p^x (1 - p)^{1 - x}$$

* mean: $p$
* variance: $p * (1-p)$


```{r}
# generate 10 bernoulli random variables with p=0.5
rbinom(10,1,0.5)
# generate 10 bernoulli random variables with p=0.2
rbinom(10,1,0.2)
```

```{r}
# density of a bernoulli random variable, p = 0.9
dbinom(1,1,0.2)
```

&nbsp;

### Binomial distribution

* sum of IID (independent and identically distributed) Bernoulli trials
* examples: number of tails in 5 coin flips, number of leads closed, ...
* parameter(s): n = number of trials, k = number of successes, p = probability of success

$$P(X = x; n,p) = 
\left(
\begin{array}{c}
  n \\ x
\end{array}
\right)
p^x(1 - p)^{n-x}$$

* mean: $np$
* variance: $np(1-p)$

```{r}
# generate 10 binomial random variables with 10 trials, p=0.2
rbinom(10,10,0.2)
```

```{r}
# generate 10 binomial random variables with 10 trials, p=0.9
rbinom(10,10,0.9)
```

```{r}
#How likely is it to obtain 4 successes in 6 trials when p=0.1?
dbinom(4,6,0.1)
```

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

# Plot various binomial distributions
n <- 10
k <- seq(0,10,1)
g1 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dbinom, args = list(size = 10, p = 0.2), geom = "bar",
                                                    fill = "blue")
g2 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dbinom, args = list(size = 10, p = 0.4), geom = "bar",
                                                    fill = "green")
g3 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dbinom, args = list(size = 10, p = 0.5), geom = "bar",
                                                    fill = "violet")
g4 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dbinom, args = list(size = 10, p = 0.9), geom = "bar",
                                                    fill = "cyan")
grid.arrange(g1,g2, g3, g4, ncol=2, nrow=2, top="Binomial distributions for different p")
```


```{r, warning=FALSE, message=FALSE}
# Corresponding cumulative distributions
n <- 10
k <- seq(0,10,1)
g1 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = pbinom, args = list(size = 10, p = 0.2), geom = "quantile",
                                                    color = "blue")
g2 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = pbinom, args = list(size = 10, p = 0.4), geom = "quantile",
                                                    color = "green")
g3 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = pbinom, args = list(size = 10, p = 0.5), geom = "quantile",
                                                    color = "violet")
g4 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = pbinom, args = list(size = 10, p = 0.9), geom = "quantile",
                                                    color = "cyan")
grid.arrange(g1,g2, g3, g4, ncol=2, nrow=2, top="Corresponding cumulative distributions")
```

### Poisson distribution

* used to model counts
* examples: number of add clicks per time interval, number of laptop crashes, ...
* with n very large and p very small, is an accurate approximation of the binomial distribution
* parameter(s): $\lambda$ = mean and variance of the distribution 

$$P(X = x; \lambda) = \frac{\lambda^x e^{-\lambda}}{x!}$$

* mean: $\lambda$
* variance: $\lambda$


```{r, warning=FALSE, message=FALSE}
n <- 10
k <- seq(0,20,1)
g1 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dpois, args = list(lambda = 1), geom = "bar",
                                                    fill = "blue")
g2 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dpois, args = list(lambda = 2), geom = "bar",
                                                    fill = "green")
g3 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dpois, args = list(lambda = 4), geom = "bar",
                                                    fill = "violet")
g4 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = dpois, args = list(lambda = 10), geom = "bar",
                                                    fill = "cyan")
grid.arrange(g1,g2, g3, g4, ncol=2, nrow=2, top="Poisson distributions for different lambda")
```

```{r, warning=FALSE, message=FALSE}
n <- 10
k <- seq(0,20,1)
g1 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = ppois, args = list(lambda = 1), geom = "quantile",
                                                    color = "blue")
g2 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = ppois, args = list(lambda = 2), geom = "quantile",
                                                    color = "green")
g3 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = ppois, args = list(lambda = 4), geom = "quantile",
                                                    color = "violet")
g4 <- ggplot(data_frame(k), aes(k)) + stat_function(fun = ppois, args = list(lambda = 10), geom = "quantile",
                                                    color = "cyan")
grid.arrange(g1,g2, g3, g4, ncol=2, nrow=2, top="Corresponding cumulative distributions")
```


Use Poisson to model rates over time:

* $t$ = total elapsed time
* expected count per unit of time: $\lambda = E[X / t]$ 
* distribution: $X \sim Poisson(\lambda t)$

Example:

* cars approaching motorway exit <fill in your favorite bottleneck location here>
* number of cars arriving per minute (on average): 12
* What is the probability that exactly 0 cars arrive in one minute? 

```{r}
dpois(0, lambda=12)
```

* Let's observe for 10 minutes in total
* What is the probability that fewer than 40 cars arrive during these 10 minutes? 

```{r}
ppois(80, lambda=12*10)
```

