---  
title: Discrete Probability Distributions
output: 
  html_document
---
  
  
## Random Variables and Probability Distributions

* A random variable models the numerical outcome of an experiment
* Random variables come in two flavours:
    + discrete: can take on only a countable number of possibilities
    + continuous: can take any value on [a subset of] the real number line 
* Discrete random variables have probability mass functions
    + every possible outcome is assigned a probability
* Continuous random variables have probability density functions
    + probability at any given _point_ is 0, can only meaningfully specify probabilities for _intervals_

## Discrete Probability Distributions


### Bernoulli Distribution

* outcome is binary with values 0 or 1
* p = probability of success (1)
* examples: coin flips,

$$P(X = x) =  p^x (1 - p)^{1 - x}$$

```{r}
# generate 10 bernoulli random variables with p=0.5
rbinom(10,1,0.5)
```

```{r}
# generate 10 bernoulli random variables with p=0.2
rbinom(10,1,0.2)
```

&nbsp;

### Binomial distribution

* sum of IID (independent and identically distributed) Bernoulli trials
* n = number of trials, k = number of successes, p = probability of success

$$P(X = x) = 
\left(
\begin{array}{c}
  n \\ x
\end{array}
\right)
p^x(1 - p)^{n-x}$$

```{r}
# generate 10 binomial random variables with 10 trials, p=0.5
rbinom(10,10,0.2)
```

```{r}
# generate 10 binomial random variables with 10 trials, p=0.9
rbinom(10,10,0.9)
```

```{r}
#How likely is it to obtain 4 success in 6 trials when p=0.1?
dbinom(4,6,0.1)
```

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

# Plot various binomial distributions
n <- 10
k <- seq(0,10,1)
p1 <- 0.1
prob1 <- dbinom(k,n,p1)
p3 <- 0.3
prob3 <- dbinom(k,n,p3)
p5 <- 0.5
prob5 <- dbinom(k,n,p5)
p9 <- 0.9
prob9 <- dbinom(k,n,p9)
g1 <- ggplot(data_frame(x = k, y1 = prob1)) + geom_bar(aes(x = k, y = prob1), stat="identity", fill="blue")
g3 <- ggplot(data_frame(x = k, y1 = prob3)) + geom_bar(aes(x = k, y = prob3), stat="identity", fill="green")
g5 <- ggplot(data_frame(x = k, y1 = prob5)) + geom_bar(aes(x = k, y = prob5), stat="identity", fill="violet")
g9 <- ggplot(data_frame(x = k, y1 = prob9)) + geom_bar(aes(x = k, y = prob9), stat="identity", fill="cyan")
grid.arrange(g1,g3,g5,g9, ncol=2, nrow=2)
```

