---  
title: Expected values
output: 
  html_document
---

## Statistical inference and expected values

In the "real world", we mostly do not have access to the "true" probability distribution.

Instead, we _estimate_ characteristics of the _population_ from a _sample_.

Concretely, we estimate _expected values_ of a population from _sample statistics_:

* population mean from sample mean
* population variance from sample variance


### Mean / expected value of a distribution

The expected mean $$E[X] = \sum_x x*p(x)$$ of a discrete probability distribution is estimated from the sample mean $$\bar X = \sum_{i=1}^n x_i  * p(x_i)$$.

The sample mean is the center of mass of the observed values.


### Example: Calculating means with R

```{r}
# roll dice 6 times
roll_dice <- sample(1:6, 6, replace = TRUE)
roll_dice
mean(roll_dice)

# roll dice 60 times
roll_dice <- sample(1:6, 60, replace = TRUE)
mean(roll_dice)

# roll dice 600 times
roll_dice <- sample(1:6, 600, replace = TRUE)
mean(roll_dice)
```


### Variance 

The variance indicates how spread out a distribution is.

Compare:

```{r, warning=FALSE, message=FALSE}
library(ggfortify)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'blue', mean=0, sd=1)
p <- ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'green', p = p, mean=0, sd=0.5)
ggdistribution(dnorm, seq(-4, 4, 0.1), colour = 'cyan', p = p, mean=0, sd=2)
```

Again, we distinguish between population variance and sample variance.

The population variance is defined as

$$Var(X) = E[(X - \mu)^2] = E[X^2] - E[X]^2$$

The sample variance is
$$S^2 = \frac{\sum_{i=1} (X_i - \bar X)^2}{n-1}$$


### Example: Calculating variances with R

```{r}
# roll dice 6 times
roll_dice <- sample(1:6, 6, replace = TRUE)
roll_dice
var(roll_dice)

# roll dice 60 times
roll_dice <- sample(1:6, 60, replace = TRUE)
var(roll_dice)

# roll dice 600 times
roll_dice <- sample(1:6, 600, replace = TRUE)
var(roll_dice)
```

### Standard deviation

The square root of the variance is called the _standard deviation_.

* In contrast to the variance, the standard deviation has the same unit as the data (and the mean)
* In R:
```{r}
roll_dice <- sample(1:6, 600, replace = TRUE)
paste0("Variance: ", var(roll_dice), ", standard deviation: ", sd(roll_dice))
```

