---
title: "probability_distributions"
output: html_document
---

## Probability distributions

### Random Variables 

* A random variable models the numerical outcome of an experiment
* Random variables come in two flavours:
    + discrete: can take on only a countable number of possibilities
    + continuous: can take any value on [a subset of] the real number line 
* Examples:
    + Die rolls are discrete random variables, where every outcome between 1 and 6 is equally likely ($P(X = x) = \frac{1}{6}$)
    + Test results are mostly regarded as continuous random variables, with an average value and a spread
 
    
### Probability mass/density function (PMF/PDF)    
* Discrete random variables have probability mass functions
    + every possible outcome is assigned a probability
* Continuous random variables have probability density functions
    + probability at any given _point_ is 0, can only meaningfully specify probabilities for _intervals_
 
  
### Cumulative distribution function (CDF)  

* The cumulative distribution function (CDF) designates the probability that the random variable is less than or equal to some value x:
$$F(x)=P(X \leq x)$$

* Examples:
    + Die rolls: $P(X \leq 3) = 0.5$
    + For a standardized IQ test, $P(X \leq 100) = 0.5$

### Quantiles

* Related to cumulative distribution functions are _quantiles_ (or _percentiles_, when multiplied by 100).
* The  $\alpha^{th}$ quantile of a distribution with distribution function $F$ is the point $x_\alpha$ so that

$$F(x_\alpha) = \alpha$$

* Examples:
    + the 0.95 quantile of a distribution is the point so that 95% of the density lies below it.
    + the 0.5 quantile is the _median_, the point such that 50% of the density lies below it
    + so, the median of the "roll dice distribution" is 3.5
* Important quantiles
    + 0.025, 0.05: often used in hypothesis testing
    + .25: first quartile, 25%
    + .5 median, 50%
    + .75 third quartile, 75%
    + 0.975, 0.95: often used in hypothesis testing


### Handling probability distributions in R

For every probability distribution, there are 4 functions in R:

* r<...>: generates a random value from the distribution
* d<...>: gives the density/mass for a value
* p<...>: gives the cumulative probability for a quantile
* q<...>: gives the corresponding value for a cumulative probability (quantile function)

Examples: 

#### Generate random variables:
```{r}
# generate one random value from a uniform distribution between 0 and 1
runif(1)
# generate 10 random values from a uniform distribution between 0 and 1
runif(10)
# generate 10 random values from a uniform distribution between -10 and 10
runif(10, min = -10, max = 10)
```

#### Get density:
```{r}
# probability density of a random value, drawn from a uniform distribution between 0 and 1
dunif(0.1)
# probability densities of 10 random values, drawn from a uniform distribution between 0 and 1
dunif(seq(0,1,.1))
# probability densities of 10 random values, drawn from a uniform distribution between -10 and 10
dunif(seq(0,1,.1),min = -10, max = 10)
```

#### Get cumulative probability:
```{r}
# cumulative probability of a random value, drawn from a uniform distribution between 0 and 1
punif(0.3)
# cumulative probability of 10 random values, drawn from a uniform distribution between 0 and 1
punif(seq(0,1,.1))
# cumulative probability of 10 random values, drawn from a uniform distribution between -10 and 10
punif(seq(0,1,.1), min = -10, max = 10)
```

#### Get value corresponding to cumulative probability:
```{r}
# value corresponding to a cumulative probability, drawn from a uniform distribution between 0 and 1
qunif(0.3)
# values corresponding to cumulative probabilities, drawn from a uniform distribution between 0 and 1
qunif(seq(0,1,.1))
# cumulativeprobability of 10 random values, drawn from a uniform distribution between -10 and 10
qunif(seq(0,1,.1), min = -10, max = 10)
```

